Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-10-07 paper/mca1.tex
AEC draft

We derive a model of concurrent computation by adding \emph{parallel
  composition} and \emph{reads-from} to \reffig{fig:seq}.  To model coherence
and synchronization, we add \emph{delay} to the rule for sequential
composition.  For \mca{} architectures, it is sufficient to encode delay in
the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
for relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.

@@

We define \PwTmca{1} by extending \refdef{def:pomset} and \reffig{fig:seq}.
The definition uses several relations over actions---$\rmatches$, $\rblocks$
and $\rdelays$---as well a distinguished set of $\sread$ actions; see
\textsection\ref{sec:actions}.
\begin{definition}
  \label{def:pwt:mca1}
  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
  with a relation $\rrfx$ such that 
  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
    \setcounter{enumi}{\value{Brf}}
  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
    is an injective relation capturing \emph{reads-from}, such that
    % \end{enumerate}
    % A pomset is a \emph{candidate} if there is an injective relation
    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
    \begin{enumerate}
      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
      % \item \label{rf-injective}
      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
      %   ${\rrfx}$ is injective,
    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
      $\aEv\le\cEv$,
    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\lt\aEv$.
    \end{enumerate}
  \end{enumerate}

  \noindent
  \label{def:pwt:mca:complete}
  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
  \begin{multicols}{2}
    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
    \item[\eqref{top-kappa}]
      \eqref{top-term}\;
      as in \refdef{def:pomset},
      \setcounter{enumi}{\value{Brf}}
    \item \label{top-rf}
      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
    \end{enumerate}
  \end{multicols}
  \bigskip

  %The semantic rules are as follows:
  \input{fig-mca1.tex}
\end{definition}
Let
\begin{math}
  \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
\end{math}
We write $\semmca{1}{}$ for the semantic function when it is unclear from context.

@@

Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
discussion of items \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2},
see \textsection\ref{sec:delay}.

>>>>

We derive a model of concurrent computation by adding
% \emph{parallel composition} and 
\emph{reads-from} to \reffig{fig:seq}.  To model coherence
and synchronization, we add \emph{delay} to the rule for sequential
composition.  For \emph{multicopy-atomic} (\mca{}) architectures, it is sufficient to encode delay in
the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
for relaxed access on \armeight{}, but requires extra synchronization for
acquiring reads.  (\emph{Lowering} is the translation of language-level
operators to machine instructions.  A lowering is \emph{optimal} if it
provides the most efficient execution possible.)

@@

We define \PwTmca{1} by extending \refdef{def:pomset} and \reffig{fig:seq}.
The definition uses several relations over actions---$\rmatches$, $\rblocks$
and $\rdelays$---as well a distinguished set of $\sread$ actions; see
\textsection\ref{sec:actions}.
\begin{definition}
  \label{def:pwt:mca1}
  The definition of \PwTmca{1} extends that of \PwT{} %(\refdef{def:pomset})
  with a relation $\rrfx$ such that 
  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
    \setcounter{enumi}{\value{Brf}}
  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
    is an injective relation capturing \emph{reads-from}, such that
    % \end{enumerate}
    % A pomset is a \emph{candidate} if there is an injective relation
    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
    \begin{enumerate}
      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
      % \item \label{rf-injective}
      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
      %   ${\rrfx}$ is injective,
    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
      $\aEv\le\cEv$,
    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\lt\aEv$.
    \end{enumerate}
  \end{enumerate}

  \noindent
  \label{def:pwt:mca:complete}
  The definition of completeness extends \refdef{def:pomset} as follows:
  \begin{multicols}{2}
    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
      \setcounter{enumi}{\value{Brf}}
    \item \label{top-rf}
      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
    \end{enumerate}
  \end{multicols}

  \noindent
  The semantic function extends \reffig{fig:seq} as follows:
  \begin{multicols}{2}
    \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
    \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
      \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
      \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
      ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$,
      \setcounter{enumi}{\value{le}}
    \item[]
      \begin{enumerate}[leftmargin=0pt]
      \item \label{seq-le-delays}
        if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.
      \end{enumerate}
    \end{enumerate}  
  \end{multicols}
  % The semantic rules are as follows:
  % \input{fig-mca1.tex}
\end{definition}
% Let
% \begin{math}
%   \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
% \end{math}
We write $\semmca{1}{}$ for the semantic function when it is unclear from context.

@@

Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
discussion of item \eqref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
see \textsection\ref{sec:delay}.

### 2021-10-07 paper/mca2.tex
in SEQ, dep respects dep1 and dep2 as needed for associativity with bimplication
AEC draft

\begin{definition}
  \label{def:pwt:mca2}
  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
  relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
  \ref{pom-rf-block} of \refdef{def:pwt:mca1}.

  A \PwTmca{2} is \emph{complete} if it satisfies
  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}---this is the same as for
  \PwTmca{1}.

  \input{fig-mca2.tex}
\end{definition}
% We write $\semmca{2}{}$ for the semantic function.
A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$.

% 
% The use of $\rextendsdef{}{}$ in \ref{if-rf-extends}
% ensures that no $\rrfx$ is introduced between events in
% $\aEvs_1\cap\aEvs_2$ when coalescing.

@@

\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. To make
  space, \reflem{lem:mca2} could move to appendix.}

With the weakening of \ref{seq-le-delays}, we must be careful not to allow
spurious pairs to be added to the $\rrfx$ relation.  Thus we add
\ref{par-rf-extends},
\ref{seq-rf-extends}, and
\ref{if-rf-extends}.
For example, \ref{if-rf-extends} ensure that
\begin{math}
  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
\end{math}
does not include 
\begin{math}
  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
        \event{a}{\DR{x}{1}}{}
        \event{b}{\DW{x}{1}}{right=of a}
        \rfint[out=165,in=15]{b}{a}
        \wki{a}{b}
      \end{tikzinlinesmall}}}
\end{math}, taking $\rrfx$ from the left and $\lt$ from the right.

% \PwTmca{2} does not enforce \ref{pom-rf-le}: $\bEv\xrfx\aEv$ may not imply
% $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from different sides of a
% sequential composition.  This means that
As a consequence of dropping \ref{pom-rf-le}, sequential $\rrfx$ must be validated during
pomset construction, rather than post-hoc.  In \textsection\ref{sec:c11}, we
show how to construct program order ($\rpox$) for complete pomsets using
phantom events ($\fmrg{}$).  Using this construction, the following lemma
gives a post-hoc verification technique for $\rrfx$.
\begin{lemma}
  \label{lem:mca2}
  If $\aPS\in\frf{\semmca{2}{\aCmd}}$ is complete, then
  for every $\bEv\xrfx\aEv$ either
  \begin{itemize}
  \item external fulfillment:
    $\bEv\lt\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
  \item internal fulfillment:
    $(\exists\bEv'\in\fmrginv{\bEv})$
    $(\exists\aEv'\in\fmrginv{\aEv})$
    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv')$
    $\labelingForm(\cEv)$ is a tautology and
    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv'\xpox\cEv\xpox\aEv'$.
  \end{itemize}
\end{lemma}
These mimic the \emph{external consistency} requirements of \armeight{}
\cite{armed}.

>>>>

\begin{definition}
  \label{def:pwt:mca2}
  The definition of \PwTmca{2} is derived from that of \PwTmca{1} by
  removing \ref{pom-rf-le} and adding the following:

  \begin{enumerate}[topsep=0pt]
  \item[{\labeltext[\textsc{p}6a]{(\textsc{p}6a)}{par-le-rf1}}]
    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\bEv\xrfx\aEv$ then $\bEv\lt\aEv$,
  \item[{\labeltext[\textsc{p}6b]{(\textsc{p}6b)}{par-le-rf2}}]
    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\aEv\xrfx\bEv$ then $\aEv\lt\bEv$,
  \item[{\labeltext[\textsc{s}6a]{(\textsc{s}6a)}{seq-le-delays-rf}}]
    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ 
    then either $\bEv\xrfx\aEv$ or $\bEv\le\aEv$,
  \item[{\labeltext[\textsc{s}6b]{(\textsc{s}6b)}{seq-le-rf-rf}}]
    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rblocks \labeling_2(\aEv)$
    then $\cEv\xrfx\aEv$ implies $\bEv\le\cEv$.
    % \\
    % if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ and $\bEv\xrfx\aEv$
    % then $\cEv\le\bEv$.
  % \item[{\labeltext[\textsc{i}6a]{(\textsc{i}6a)}{if-le-rf}}]
  %   ${\rrfx}\subseteq {\rrfx_1}\cup {\rrfx_2}$
  \end{enumerate}
  % The definition of \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
  % relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
  % \ref{pom-rf-block} of \refdef{def:pwt:mca1}.

  % A \PwTmca{2} is \emph{complete} if it satisfies
  % \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}---this is the same as for
  % \PwTmca{1}.

  %\input{fig-mca2.tex}
\end{definition}
% We write $\semmca{2}{}$ for the semantic function.
A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$.

% 
% The use of $\rextendsdef{}{}$ in \ref{if-rf-project}
% ensures that no $\rrfx$ is introduced between events in
% $\aEvs_1\cap\aEvs_2$ when coalescing.

@@

\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. Perhaps
  move \reflem{lem:mca2} to appendix.}

With the weakening of \ref{seq-le-delays}, we must be careful not to allow
spurious pairs to be added to the $\rrfx$ relation.  For example, %\ref{if-rf-project} ensures that
\begin{math}
  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
\end{math}
should not include 
\begin{math}
  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
        \event{a}{\DR{x}{1}}{}
        \event{b}{\DW{x}{1}}{right=of a}
        \rfint[out=165,in=15]{b}{a}
        \wki{a}{b}
      \end{tikzinlinesmall}}}
\end{math}, taking $\rrfx$ from the left and $\lt$ from the right.
The use of $\rextends{}{}$ ensures this.

% \PwTmca{2} does not enforce \ref{pom-rf-le}: $\bEv\xrfx\aEv$ may not imply
% $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from different sides of a
% sequential composition.  This means that
As a consequence of dropping \ref{pom-rf-le}, sequential $\rrfx$ must be validated during
pomset construction, rather than post-hoc.  In \textsection\ref{sec:c11}, we
show how to construct program order ($\rpox$) for complete pomsets using
phantom events ($\fmrg{}$).  Using this construction, the following lemma
gives a post-hoc verification technique for $\rrfx$.  Let $\fmrginv{}$ be the
inverse of $\fmrg{}$.
\begin{lemma}
  \label{lem:mca2}
  If $\aPS\in\frf{\semmca{2}{\aCmd}}$ is complete, then
  for every $\bEv\xrfx\aEv$ either
  \begin{itemize}
  \item external fulfillment:
    $\bEv\lt\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
  \item internal fulfillment:
    $(\exists\bEv'\in\fmrginv{\bEv})$
    $(\exists\aEv'\in\fmrginv{\aEv})$
    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv)$
    $\labelingForm(\cEv)$ is a tautology and
    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv'\xpox\cEv\xpox\aEv'$.
  \end{itemize}
\end{lemma}
These mimic the \emph{external consistency} requirements of \armeight{}
\cite{armed}.
